Basics of Hardware Cache Optimization

### 2017-05-28 tex/basic-concepts.tex
Add awesome plot, expand content, tweak stuff
Minor update
Revise, tweak, expand

\subsection{Cache Line} % or Cache Block
% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
%
% > On x86/x64, cache line is 64 bytes for many years now.
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
%      -- Drepper, p. 15
% > It is not possible for a cache to hold partial cache lines.
%      -- Drepper, p. 16
\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
{scott-meyers-talk}.%
\footnote{%
% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
% > cache 256 KB in size, with 128-byte cache blocks.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
% TODO: what about the block size of main memory?  Should it be the same?
Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
line}~\cite[p.~11]{pentium4}.}
\begin{comment}
   \multiplefootnoteseparator%
   % See <https://tex.stackexchange.com/a/71015>.
   \footnote{This can also be checked on the command line:
   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
\end{comment}
% > It means that as soon as you've accessed any single byte in a cache line, all the
% > other 63 bytes are already in L1 cache
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
\alts{%
   This means accessing a single uncached 32-bit integer entails loading another 60
   adjacent bytes.,
   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
   well.},
   {When, for example, accessing a single byte that isn't already cached, another 63
   adjacent bytes will be loaded.}
}
% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.

My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

@@

\begin{listing}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{listing}

@@

\subsection{Prefetching}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching
\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

>>>>

\subsection{Cache Line} % or Cache Block
% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
%
% > On x86/x64, cache line is 64 bytes for many years now.
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
%      -- Drepper, p. 15
% > It is not possible for a cache to hold partial cache lines.
%      -- Drepper, p. 16
% > A cache line is the "unit" of data you transfer to a cache.
%      -- http://www.cs.umd.edu/class/sum2003/cmsc311/Notes/Memory/introCache.html
\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
{scott-meyers-talk}.%
\footnote{%
% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
% > cache 256 KB in size, with 128-byte cache blocks.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
% TODO: what about the block size of main memory?  Should it be the same?
Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
line}~\cite[p.~11]{pentium4}.}
\begin{comment}
   \multiplefootnoteseparator%
   % See <https://tex.stackexchange.com/a/71015>.
   \footnote{This can also be checked on the command line:
   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
\end{comment}
% > It means that as soon as you've accessed any single byte in a cache line, all the
% > other 63 bytes are already in L1 cache
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
\alts{%
   This means accessing a single uncached 32-bit integer entails loading another 60
   adjacent bytes.,
   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
   well.},
   {When, for example, accessing a single byte that isn't already cached, another 63
   adjacent bytes will be loaded.}
}
% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.

My E-450 CPU is no exception and both of its data caches have 64-byte cache lines.%
\footnote{See \cref{app:cpuinfo}.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
\begin{center}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \captionof{listing}{%
      Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{center}
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

@@

\begin{comment}
   \begin{listing}
      \inputminted[firstline=27]{c}{line-size/line-size.c}
      \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
      \label{lst:line-size}
   \end{listing}
\end{comment}

@@

% Talk about how the term cache line is commonly conflated to mean an appropriately sized
% and *aligned* block of main memory that can be loaded into a cache line.  This is how it
% works, right?  Or can we load blocks starting at arbitrary addresses into a cache line?
% One could even imagine loading memory that isn't even contiguous.  Most sources don't
% seem to clear any of this up.  See <https://stackoverflow.com/q/3928995> ("How do cache
% lines work?")
\begin{comment}
   The term cache line is also used to refer to \alts{identically, appropriately} sized
   blocks in main memory.
   % that may be loaded into cache.
   These blocks are fixed:
   % I.e., each byte in main memory falls exactly into one cache line.
   % The boundaries

   Cache lines in main memory are fixed.
   % Data is not loaded starting from arbitrary addresses, but only from addresses that
   % are multiples of the cache line size.
   Data blocks that are loaded don't start at arbitrary addresses, but at multiples of the
   cache line size.
\end{comment}
% See <https://en.wikipedia.org/wiki/Partition_of_a_set>.  XXX: I feel like there's some
% kind of conflation of ideas going on here.
Both cache and main memory can be thought of as being partitioned%
% \footnote{%
%    In the set-theoretic sense
% }
\ (in the set-theoretic\x{al}\ sense) % https://en.wiktionary.org/wiki/set-theoretic
into \alts[3]{cache line-sized blocks, blocks of that size, cache lines}.  \alts[2]{{That
is, d}, D}ata is \alts{not, neither} \alts{read or written, loaded} starting from
arbitrary main memory addresses,
% ...nor is it loaded into arbitrarily aligned blocks in cache.
but only from addresses that are multiples of the cache line size.
% The starting address will by a multiple of the cache line size.

% TODO: this probably has some implications about aligning data.

% TODO: maybe talk about the 64-bit bus width and burst mode as well.  And maybe about how
% a write requires a read if we only write part of a cache line and the optimization of
% adding dummy writes (I think Andrei talked about this).
%
% [1]: https://stackoverflow.com/q/39182060 "Why isn't there a data bus which is as wide
%      as the cache line size?"

\subsection{Prefetching}
\label{sec:prefetch}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching
\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

### 2017-05-28 tex/intro.tex
Revise, tweak, expand

% What are hardware caches?  Why caches?
A hardware cache is a \alts{comparatively, relatively} fast and small physical memory.  It
stores a subset of the data present in slower, larger memory that is expected to be used
again soon.  The purpose of this additional memory is to reduce the number of accesses to
the underlying slower storage.

@@

The most ubiquitous example for hardware caches \alts{is the hierarchy, are the various
levels (most commonly 2 or 3)} of CPU caches that are found on almost all present-day
CPUs.  They are designated L1 cache, L2 cache, and so on, with L1 being the fastest and
smallest level.  The underlying storage for CPU caches is the main memory.

>>>>

% What are hardware caches?  Why caches?
A hardware cache is a \alts{comparatively, relatively} fast and small physical memory.  It
stores a subset of the data present in slower, larger \alts{storage, memory} that is
expected to be used again soon.  The purpose of this additional memory is to reduce the
number of accesses to the underlying slower storage.

@@

The most ubiquitous example for hardware caches \alts{is the hierarchy, are the various
levels (most commonly 2 or 3)} of CPU caches that are found on almost all present-day
CPUs.  They are designated \gls{l1} cache, \gls{l2} cache, and so on, with \gls{l1} being
the fastest and smallest level.  The underlying storage for CPU caches is the main memory.

### 2017-05-28 tex/types.tex
Revise, tweak, expand

% > [Instruction] cache is much less problematic than the data cache.
%      -- Drepper, p. 31
The biggest target for optimizations is the data cache.  \say{[Instruction] cache is much
less problematic}~\cite[31]{drepper2007} and optimizations for data and instruction cache
tend to improve \gls{tlb} usage as well%
~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m53s}{11:53}]{scott-meyers-talk}.

@@

@@

% \Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
% when not having defined \mintinline{text}{BASELINE}.
\Cref{fig:access-times} shows the extra CPU cycles used by \cref{lst:access-times}
\alts[2]{above, in addition to, compared to} the \mintinline{text}{BASELINE} version for
different array sizes.
That is, only the cycles used by the main loop are \alts{given, counted}, not those for
initialization.  I divided by \mintinline{text}{N} to get the cycles spent \alts{per, on
each} loop iteration.

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{The numerical results are shown in \cref{tab:access-times}.}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets evicted.  The
first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} hits and an \gls{l2} access time of around 25 cycles.

The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
increase when the array size matches that of the \gls{l2}, 512 KiB, is \alts{greater, more
striking} than for the \gls{l1d} before; possibly because \gls{l2} is a unified cache that
also holds instructions.  Eventually, more and more accesses go to main memory,
\alts{causing, incurring} delays of up to 200 cycles~\cite[cf.][17]{drepper2007}.

>>>>

% > [Instruction] cache is much less problematic than the data cache.
%      -- Drepper, p. 31
% TODO: move this paragraph?
The biggest target for optimizations is the data cache.  \say{[Instruction] cache is much
less problematic}~\cite[31]{drepper2007} and optimizations for data and instruction cache
tend to improve \gls{tlb} usage as well%
~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m53s}{11:53}]{scott-meyers-talk}.

@@

\begin{comment}
   We will see why this measurement was conducted with random instead of sequential
   accesses in \cref{sec:prefetch}.

   \Cref{sec:prefetch} \alts{explains, will explain} the reason for \x{going through the
   trouble of} \alts{using, conducting} random accesses instead of just reading the array
   sequentially\x{for this measurement}.
\end{comment}
\alts{We use random accesses because,
      The reason for using random accesses is that}
the CPU will detect and optimize sequential access by a technique called
\emph{prefetching} discussed in \cref{sec:prefetch}, which would prevent us from
\alts{determining, assessing} access times.

@@

% \Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
% when not having defined \mintinline{text}{BASELINE}.
\Cref{fig:access-times} shows the extra CPU cycles used by \cref{lst:access-times}
\alts{in addition to, above, compared to, relative to} the \mintinline{text}{BASELINE}
version for different array sizes.
That is, only the cycles used by the main loop are \alts{given, counted}, not those for
initialization.  I divided by \mintinline{text}{N} to get the cycles spent \alts{per, on
each} loop iteration.

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{The numerical results are shown in \vref{tab:access-times}.}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets \emph{evicted}.
The first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} hits and an \gls{l2} access time \x{of} around 25 cycles.

The values from 512 KiB (the size of the \gls{l2}) to 128 MiB \alts{exhibit, follow} a
similar pattern.  The relative increase when the array size matches that of the \gls{l2}
is \alts{greater, more striking} than for the \gls{l1d} before; possibly because \gls{l2}
is a unified cache that also holds instructions.  Eventually, more and more accesses go to
main memory, \alts{causing, incurring} delays of up to 200
cycles~\cite[cf.][\pno~17, figure 3.4]{drepper2007}.

% FOLDOC has an entry on 'working set'.
\alts[2]{{Evidently,}, The data suggests that} keeping the \emph{working set} a process
uses during a time interval small can yield dramatic performance improvements.
